---
title: "Simple example"
author: "Michael Bedward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

In this example we will:

* create a set of randomly positioned, overlapping circles of various sizes;
* use the `circleLayout` function to find a non-overlapping arrangement;
* use `ggplot` to draw the before and after arrangements.

Firstly lets create some random circles, positioned within the central portion
of a bounding square, with smaller circles being more common than larger ones.

```{r}

ncircles <- 200
limits <- c(-50, 50)
inset <- diff(limits) / 3
rmax <- 20

xyr <- data.frame(
  x = runif(ncircles, min(limits) + inset, max(limits) - inset),
  y = runif(ncircles, min(limits) + inset, max(limits) - inset),
  r = rbeta(ncircles, 1, 10) * rmax)
```

Next, we use the `circleLayout` function to try to find a non-overlapping
arragement, allowing the circles to occupy any part of the bounding square.
The returned value is a list with elements for the layout and the number
of iterations performed.

```{r}
library(packcircles)

res <- circleLayout(xyr, limits, limits, maxiter = 1000)
cat(res$niter, "iterations performed")
```

Now we can use ggplot to draw the before and after layouts.

```{r, fig.width=7, fig.height=4}
library(ggplot2)
library(gridExtra)

## plot data for the `before` layout
dat.before <- circlePlotData(xyr)

## plot dta for the `after` layout returned by circleLayout
dat.after <- circlePlotData(res$layout)

doPlot <- function(dat, title)
  ggplot(dat) + 
  geom_polygon(aes(x, y, group=id), colour="brown", fill="burlywood", alpha=0.3) +
  coord_equal(xlim=limits, ylim=limits) +
  theme_bw() +
  theme(axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank()) +
  labs(title=title)

grid.arrange(
  doPlot(dat.before, "before"),
  doPlot(dat.after, "after"),
  nrow=1)
```

