--- 
title: "Progressive circle packing" 
author: "Michael Bedward" 
date: "`r Sys.Date()`" 
output: rmarkdown::html_vignette 
vignette: > 
  %\VignetteIndexEntry{Progressive circle packing}
  %\VignetteEngine{knitr::rmarkdown} 
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}

set.seed(42)

library(packcircles) 
library(ggplot2)

t <- theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.text=element_blank(),
        axis.ticks=element_blank(), 
        axis.title=element_blank())

theme_set(t)

```

The function `circleProgressiveLayout` arranges a set of circles
deterministically. The first two circles are placed to the left and right of the
origin respectively. Subsequent circles are placed so that each:

  * is externally tangent to two previously placed circles; 
  * does not overlap another circle; 
  * and is as close as possible to the origin.

The algorithm was described in the paper: *Visualization of large hierarchical
data by circle packing* by Weixin Wang, Hui Wang, Guozhong Dai, and Hongan Wang.
Published in *Proceedings of the SIGCHI Conference on Human Factors in Computing
Systems*, 2006, pp. 517-520. [ACM](https://dl.acm.org/citation.cfm?id=1124851).

The implementation in this package is based on a [version written in C by Peter 
Menzel](https://github.com/pmenzel/packCircles).

The algorithm is very efficient and this, combined with the implementation in Rcpp, means arrangements for large numbers of circles can be found quickly.


## First example

We begin by arranging 10 circles of various sizes. First we pass a
vector of circle areas to the `circleProgressiveLayout` function. It returns a 
data frame of centre coordinates and radii.

```{r}

areas <- c(20, 10, 40, rep(5, 7))

# Generate the layout 
packing <- circleProgressiveLayout(areas) 

head( round(packing, 2) )

```

Next we derive a data frame of circle vertices for plotting using the 
`circleLayoutVertices` function, and then use ggplot to display the layout, labelling 
the circles to show their order of placement:

```{r fig.width=4, fig.height=4} 

dat.gg <- circleLayoutVertices(packing, npoints=50)

ggplot(data = dat.gg) + geom_polygon(aes(x, y, group = id), colour = "black", 
fill = "grey90", alpha = 0.7, show.legend = FALSE) +

geom_text(data = packing, aes(x, y), label = 1:nrow(packing)) +

coord_equal()

```

By default, `circleProgressiveLayout` takes the input sizes to be circle areas. 
If instead you have input radii, add `sizetype = "radius"` to the arguments when calling the 
function.


## Layouts are order dependent

Re-ordering the input sizes will generally produce a different layout unless the sizes are uniform. Here we repeatedly shuffle the area values used above and generate a new layout each time.

```{r fig.width=7, fig.height=5}

ncircles <- length(areas) 
nreps <- 6

packings <- lapply(
  1:nreps, 
  function(i) { 
    x <- sample(areas, ncircles) 
    circleProgressiveLayout(x) 
  })

packings <- do.call(rbind, packings)

npts <- 50 
dat.gg <- circleLayoutVertices(packings, npoints = npts) 

dat.gg$rep <- rep(1:nreps, each = ncircles * (npts+1))


ggplot(data = dat.gg, aes(x, y)) + 
  geom_polygon(aes(group = id), 
               colour = "black", fill = "grey90") +

  coord_equal() +

  facet_wrap(~ rep, nrow = 2)

```


We can use this ordering effect to create some circle art...


```{r fig.width=7, fig.height=4}

areas <- 1:1000

# area: small to big
packing1 <- circleProgressiveLayout(areas)
dat1 <- circleLayoutVertices(packing1)

# area: big to small
packing2 <- circleProgressiveLayout( rev(areas) ) 
dat2 <- circleLayoutVertices(packing2)

dat <- rbind( 
  cbind(dat1, set = 1), 
  cbind(dat2, set = 2) )

ggplot(data = dat, aes(x, y)) + 
  geom_polygon(aes(group = id, fill = -id), 
               colour = "black", show.legend = FALSE) +
  
  scale_fill_distiller(palette = "RdGy") +
  
  coord_equal() +
  
  facet_wrap(~set, 
             labeller = as_labeller(
               c('1' = "small circles first", 
                 '2' = "big circles first"))
             )

```


## More detailed layout display

The package includes an example data set of the abundance of different types of bacteria measured in a study of biofilms. Columns are value (abundance), display colour and label (bacterial taxon).

```{r}

data("bacteria")
head(bacteria)

```

The following example shows how to display the abundance values as circles filled with the specified colours. It relies on the fact that the `id` column in the output of `circleLayoutVertices` maps to the row number of the input data.

Note: the y-axis is reversed so that the layour is rendered similarly to the example [here](https://github.com/pmenzel/packCircles).

```{r fig.width=5, fig.height=5, fig.align='center'}

packing <- circleProgressiveLayout(bacteria)

dat.gg <- circleLayoutVertices(packing)

ggplot(data = dat.gg) +
  geom_polygon(aes(x, y, group = id, fill = factor(id)), 
               colour = "black",
               show.legend = FALSE) +
  
  scale_fill_manual(values = bacteria$colour) +
  
  scale_y_reverse() +
  
  coord_equal()

```


As a further flourish, we can make the plot interactive so that the name of the bacterial taxon is displayed when the mouse cursor hovers a circle. Here we use the `ggiraph` package to accomplish this.

```{r fig.width=5, fig.height=5}

library(ggiraph)

gg <- ggplot(data = dat.gg) +
  
  geom_polygon_interactive(
    aes(x, y, group = id, fill = factor(id),
        tooltip = bacteria$label[id], data_id = id), 
    colour = "black",
    show.legend = FALSE) +
  
  scale_fill_manual(values = bacteria$colour) +
  
  scale_y_reverse() +
  
  labs(title = "Hover over circle to display taxon name") +

  coord_equal()

ggiraph(ggobj = gg, width_svg = 5, height_svg = 5)

```

